---
title: "Data Wrangling in R"
author: "Haseeb Raza (Microbiologist | Data Analyst | Bioinformatician)"
date: "2024-09-03"
output:
  html_document: 
    highlight: zenburn
  pdf_document: 
    highlight: zenburn
---
# Activating Packages

```{r required_packages, message=FALSE, warning=FALSE, paged.print=FALSE}
library(readxl)
library(tidyverse)
```

In the above code chunk we have activated the packages required for data wrangling/transformation.

# Importing dataset from excel 
```{r importing_data, message=FALSE, warning=FALSE, paged.print=FALSE}
library(readxl)
df <- read_excel("_Participatory Epidemiology of Ticks and Tick-Borne Piroplasmosis along with their Therapeutic and Control Measures_. (Responses).xlsx", 
    range = "A1:BS1236", col_types = c("date", 
        "text", "text", "text", "numeric", 
        "text", "text", "text", "text", "text", 
        "text", "text", "text", "text", "text", 
        "numeric", "text", "text", "text", 
        "text", "text", "text", "text", "text", 
        "text", "text", "text", "text", "text", 
        "text", "text", "text", "text", "text", 
        "text", "text", "text", "text", "text", 
        "text", "text", "text", "text", "text", 
        "text", "text", "text", "text", "text", 
        "text", "text", "text", "text", "text", 
        "text", "text", "text", "text", "text", 
        "text", "text", "text", "text", "text", 
        "text", "text", "text", "text", "text", 
        "text", "text"))
```


# Understanding the data structure 
```{r tibble_df, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
as.tibble(df)
# head(df)
# tail(df)
# glimpse(df)
# View(df)
# summary(df)
```

# Subsetting Data on the basis of Variables using select() Function

```{r selecting_variables, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
# df <- df[,-c(1,2,3)]
# select(df, "Gender" , "Education Status") %>% head()

df_1 <- select(df, 'Gender', 
        'Education Status', 
        'Occupation type', 
        'Socioeconomic Status',
        'Name of District', 
        'Name of Tehsil', 
        'Living Area', 
        'Based on your observations, how would you rank the ticks in the matter of frequency', 
       'In which host you have observed ticks?',
        'If it was animal host then select the host species in which ticks were observed',
        'Animal Gender', 
        'Health Status of animal in which ticks are observed (select all possible)', 
        'What is your level of knowledge about ticks?', 
        'How frequently you observe the ticks?', 
        'Did you find any impact of these ticks on animal health?', 
        'What is the average number of ticks you have observed last year?', 
        'Hygienic Measures of the area (select all possible)',
        'How frequently you have observed theileriosis in your animals?',
        'How frequently you have observed babesiosis in animals?') 

head(df)
```


# Renaming Data Columns with rename() Function

```{r renaming_col, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
df_1 <- df_1 %>% rename("gender" = "Gender", 
                "educational_status" = "Education Status",
                "ocupation_type" = "Occupation type",
                "sec" = "Socioeconomic Status",
                "dist_name" = "Name of District",
                "tehsil_name" = "Name of Tehsil", 
                "living_area" = "Living Area",
                "tick_freq" = "Based on your observations, how would you rank the ticks in the matter of frequency",
                "host_animal" = "In which host you have observed ticks?" ,
                "host_specie" = "If it was animal host then select the host species in which ticks were observed", 
                "animal_gender" = "Animal Gender", 
                "host_health" =  "Health Status of animal in which ticks are observed (select all possible)",
                "knowledge" = "What is your level of knowledge about ticks?", 
                "obs_freq" = "How frequently you observe the ticks?", 
                "impact_ani_health" = "Did you find any impact of these ticks on animal health?", 
                "ave_tick_ly" = "What is the average number of ticks you have observed last year?",
                "hm" = "Hygienic Measures of the area (select all possible)" , 
                "theileriosis_freq"= "How frequently you have observed theileriosis in your animals?", 
                "babesiosis_freq" =  "How frequently you have observed babesiosis in animals?") %>% print()
```

# Recoding Data Columns with mutate() & recode() Functions

```{r recod_var, message=FALSE, warning=FALSE, paged.print=FALSE}
df_1 %>%  
  mutate(dist_name = recode(dist_name, "Rajan pur"= "Rajanpur" , "Rajan 6" = "Rajanpur", "faisalabad" = "Faisalabad", "Fsd" = "Faisalabad"  )  ) %>% 
  drop_na(dist_name) %>% 
  summarise(unique(dist_name))




df_1 <- df_1 %>%  
  mutate(dist_name = recode(dist_name, "Rajan pur"= "Rajanpur" , "Rajan 6" = "Rajanpur", "faisalabad" = "Faisalabad", "Fsd" = "Faisalabad"  )  ) %>% 
  drop_na(dist_name) 
```


# Subsetting Data on the basis of Rows using filter() Function

```{r filtering_rows, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
glimpse(df_1)
filter(df_1, dist_name == "Rajanpur" | dist_name == "Rajan pur" ) %>% 
  print()

filter(df_1, dist_name %in% c("Rajanpur","Rajan pur", "Rajan 6" )) %>%  
  print()


```


# Re-Transformation of Data 

```{r pivot_longer, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)



 df_test <- df %>% select(., 'Timestamp', 
              'Gender', 
        'Education Status', 
        'Occupation type', 
        'Socioeconomic Status',
        'Name of District', 
        'Name of Tehsil', 
        'Living Area', 
        'Based on your observations, how would you rank the ticks in the matter of frequency', 
       'In which host you have observed ticks?',
        'If it was animal host then select the host species in which ticks were observed',
        'Animal Gender', 
        'Health Status of animal in which ticks are observed (select all possible)', 
        'What is your level of knowledge about ticks?', 
        'How frequently you observe the ticks?', 
        'Did you find any impact of these ticks on animal health?', 
        'What is the average number of ticks you have observed last year?', 
        'Hygienic Measures of the area (select all possible)',
        'How frequently you have observed theileriosis in your animals?',
        'How frequently you have observed babesiosis in animals?')  %>%  rename("date"= "Timestamp",
                                                                                "gender" = "Gender", 
                "educational_status" = "Education Status",
                "ocupation_type" = "Occupation type",
                "sec" = "Socioeconomic Status",
                "dist_name" = "Name of District",
                "tehsil_name" = "Name of Tehsil", 
                "living_area" = "Living Area",
                "tick_freq" = "Based on your observations, how would you rank the ticks in the matter of frequency",
                "host_animal" = "In which host you have observed ticks?" ,
                "host_specie" = "If it was animal host then select the host species in which ticks were observed", 
                "animal_gender" = "Animal Gender", 
                "host_health" =  "Health Status of animal in which ticks are observed (select all possible)",
                "knowledge" = "What is your level of knowledge about ticks?", 
                "obs_freq" = "How frequently you observe the ticks?", 
                "impact_ani_health" = "Did you find any impact of these ticks on animal health?", 
                "ave_tick_ly" = "What is the average number of ticks you have observed last year?",
                "hm" = "Hygienic Measures of the area (select all possible)" , 
                "theileriosis_freq"= "How frequently you have observed theileriosis in your animals?", 
                "babesiosis_freq" =  "How frequently you have observed babesiosis in animals?") %>% 
  mutate(dist_name = recode(dist_name, "Rajan pur"= "Rajanpur" , "Rajan 6" = "Rajanpur", "faisalabad" = "Faisalabad", "Fsd" = "Faisalabad"  )  )
 
 
 #  %>% pivot_longer(cols=c("theileriosis_freq","babesiosis_freq"), names_to="observation", values_to = "frequency")
 
```


# Changing Nature of a variable from "dttm" to "date"

```{r dttm_to_date, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)

# df$Timestamp <- as.POSIXct(df$Timestamp)

df_test$date <- as.Date(df_test$date)
```


```{r str_df_test, message=FALSE, warning=FALSE, paged.print=FALSE}
str(df_test)
```


# Statistical Analysis of Data

```{r summary_df_test, message=FALSE, warning=FALSE, paged.print=FALSE}
summary(df_test)
```

## Distribution of Data Variables

Example to understand a "for loop" making histograms

```{r normality_check, message=FALSE, warning=FALSE, paged.print=FALSE}


set.seed(123) # For reproducibility
df <- data.frame(
  var1 = rnorm(100, mean = 50, sd = 10),
  var2 = rnorm(100, mean = 30, sd = 5),
  var3 = rnorm(100, mean = 70, sd = 15)
)


for (col in names(df)) {
  # Generate histogram
  hist(df[[col]], main = paste("Histogram of", col), xlab = col, col = "skyblue", border = "white")
}


```

*Normality of "tick_freq"*

```{r normality_check_2, message=FALSE, warning=FALSE, paged.print=FALSE}
hist(df_test$tick_freq, xlab = "Frequency of Ticks", col = "red", border = "orange")


# qqnorm(df_test$tick_freq, main = "QQ Plot of Frquency of Ticks", pch = 19, col = "blue", cex = 1.2, xlab = "Theoretical Quantiles", ylab = "Sample Quantiles") 


shapiro.test(df_test$tick_freq)

# data shows abnormal distribution in each test 
# sqrt() transformation of data 

#df_log <- df_test %>%
  #mutate(
   # log_tick_freq = log(tick_freq)
 # )

# shapiro.test(df_log$log_tick_freq)
# hist(df_log$log_tick_freq,xlab = "Frequency of Ticks", col = "red", border = "orange" )
remove(df_log)
```


## Composition of Data Variable 

```{r comp_check, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
library(car)

leveneTest(df_test$tick_freq ~ df_test$dist_name , data = df_test)

```
```{r comp_check_2, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
library(car)

leveneTest(df_test$tick_freq ~ df_test$sec , )
```


```{r comp_check_3, message=FALSE, warning=FALSE, paged.print=FALSE}


library(tidyverse)
library(car)

leveneTest(df_test$tick_freq ~ df_test$living_area)
```
```{r comp_check_4, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
library(car)

leveneTest(df_test$tick_freq ~ df_test$impact_ani_health)

leveneTest(df_test$tick_freq ~ df_test$gender)

leveneTest(df_test$tick_freq ~ df_test$educational_status)

leveneTest(df_test$tick_freq ~ df_test$tehsil_name)

leveneTest(df_test$tick_freq ~ df_test$host_animal)

leveneTest(df_test$tick_freq ~ df_test$host_specie)

leveneTest(df_test$tick_freq ~ df_test$animal_gender)

leveneTest(df_test$tick_freq ~ df_test$host_health)

leveneTest(df_test$tick_freq ~ df_test$knowledge)

leveneTest(df_test$tick_freq ~ df_test$obs_freq)

leveneTest(df_test$tick_freq ~ df_test$ave_tick_ly)

leveneTest(df_test$tick_freq ~ df_test$hm)

leveneTest(df_test$tick_freq ~ df_test$theileriosis_freq)

leveneTest(df_test$tick_freq ~ df_test$babesiosis_freq)


```

## Relationship between Data Variables 

```{r recoding_into_numbers, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)


# recoding columns for better analysis
 #df_recode <- df_test %>%
  filter(!is.na(df_test$theileriosis_freq)) %>%  
  mutate(theileriosis_freq = recode(theileriosis_freq,
                                    "1" = "Less frequent",
                                    "2" =  "Seldom",
                                    "3" = "Highly frequent"))


 #df_recode <- df_recode %>%
  filter(!is.na(babesiosis_freq)) %>%  
  mutate(babesiosis_freq = recode(babesiosis_freq,
                                    1 = "Less frequent",
                                    2 =  "Seldom",
                                    3 = "Highly frequent"))
 
 
 
 # df_recode <- df_recode %>%
  filter(!is.na(ave_tick_ly)) %>%  
  mutate(ave_tick_ly = recode(ave_tick_ly,
                                    "Above 50" = "50-60",
                                    "up to 10" = "1-10"))
  
  
  
 df_recode <-  df_recode %>% 
  separate(ave_tick_ly , into=c("min", "max"), sep="-", convert=TRUE) %>%
	mutate(midpoint = (min+max)/2)

    
library(ggplot2)

ggplot(df_recode) +
 aes(x = theileriosis_freq, y = babesiosis_freq, fill = tick_freq) +
 geom_tile() +
 scale_fill_distiller(palette = "YlOrRd", direction = 1) +
 theme_minimal() +
 facet_wrap(vars(dist_name))

 
```

Filtering data to get rows of  "Narowal" , "Rajanpur" , and "Toba Tek Singh".

```{r subsetting_three_dist, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)

df_recode <-   df_recode %>% filter(dist_name %in% c("Rajanpur","Narowal", "Toba Tek Singh" ))

```


Making visualization to show the relationship between the variables. 

```{r relationship_plotting, message=FALSE, warning=FALSE, paged.print=FALSE}
library(readxl)
library(tidyverse)
library(ggplot2)


library(ggplot2)

ggplot(df_recode) +
 aes(x = sec, fill = educational_status) +
 geom_bar() +
 scale_fill_viridis_d(option = "viridis", 
 direction = 1) +
 labs(x = "Socio Economic Status") +
 coord_polar(theta = "y") +
 theme_minimal() +
 facet_wrap(vars(dist_name))

library(ggplot2)

ggplot(df_recode) +
 aes(x = living_area, fill = hm) +
 geom_bar() +
 scale_fill_brewer(palette = "Paired", 
 direction = 1) +
 theme_classic() +
 theme(legend.position = "none", axis.text.y = element_text(size = 14L, 
 vjust = 0.1), axis.text.x = element_text(size = 14L, angle = 45L, hjust = 0.9)) +
 facet_wrap(vars(dist_name))

ggplot(df_recode) +
 aes(x = living_area, fill = hm) +
 geom_bar() +
 scale_fill_brewer(palette = "Paired", 
 direction = 1) +
 labs(x = "Living Area", y = "Counts") +
 theme_classic() +
 theme(legend.position = "none", 
 axis.title.y = element_text(size = 16L), axis.title.x = element_text(size = 16L, hjust = 1), axis.text.y = element_text(size = 14L, 
 vjust = 0.1), axis.text.x = element_text(size = 14L, angle = 45L, hjust = 0.9)) +
 facet_wrap(vars(dist_name))






```





```{r relationship_plotting_2, message=FALSE, warning=FALSE, paged.print=FALSE}
library(readxl)
library(tidyverse)
library(ggplot2)
#library(esquisse)

#esquisser()

ggplot(df_recode) +
  aes(
    x = theileriosis_freq,
    y = babesiosis_freq,
    fill = tick_freq
  ) +
  geom_tile() +
  scale_fill_distiller(palette = "YlOrRd", direction = 1) +
  theme_minimal() +
  facet_wrap(vars(dist_name))



```



